\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-09-20 doc/paper-jss/sdmTMB-paper.Rnw
Edit

A basic GLMM with spatial Gaussian random fields (GRFs) can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s}}] &= \mu_{\bm{s}},\\
\mu_{\bm{s}} &=
g^{-1} \left( \eta_{\bm{s}} \right),\\
\eta_{\bm{s}} &= \bm{X}_{\bm{s}} \bm{\beta} + \omega_{\bm{s}},\\
\boldsymbol{\omega} &\sim \operatorname{MVNormal}
  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\
\end{aligned}
\]
where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in
space $\bm{s}$ is equal to mean $\mu_s$. That mean is the result of an inverse
link function $g^{-1}$ applied to a linear predictor $\eta$. In this case, that
linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$
multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial
GRF $\omega_{\bm{s}}$. This spatial GRF represents the effect of latent spatial
variables that are not otherwise accounted for in the model. Alternatively,
$\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
More simply, $\omega_{\bm{s}}$ represents a ``wiggly'' surface with an expected
value of zero that is added to the expectation in link space.

The vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) is assumed drawn from
a multivariate normal distribution with covariance matrix
$\boldsymbol{\Sigma}_\omega$ constrained by some covariance function that
defines the rate that spatial covariance decays with distance. Various
covariance functions are possible, but a popular and flexible function is the
Mat\'ern (REF) (FIGURE TODO). If we define $\lVert h \rVert$ as the Euclidean
distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the
Mat\'ern covariance $\Phi$ as
\[
\Phi\left( \bm{s}_j, \bm{s}_k \right) =
  \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \right).
\]
The parameter $\tau^2$ controls the spatial variance, $\Gamma$ represents the
Gamma function, $K_\nu$ represents the modified Bessel function of the second
kind, and $\kappa$ represents the spatial decorrelation rate. The parameter
$\nu$ controls the smoothness of the covariance function, but can be fixed at
$\nu = 1$ to take advantage of the SPDE \citep{lindgren2011}. The parameters
$\kappa$ and $\tau$ can be converted to two useful values: (1) correlation range
as $\textrm{range} = \sqrt{8} / \kappa$ and (2) marginal standard deviation as
$\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
The range is the distance that correlation decays to $\approx 0.13$.

>>>>

A basic GLMM with spatial Gaussian random fields (GRFs) can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s}}] &= \mu_{\bm{s}},\\
\mu_{\bm{s}} &=
g^{-1} \left( \eta_{\bm{s}} \right),\\
\eta_{\bm{s}} &= \bm{X}_{\bm{s}} \bm{\beta} + \omega_{\bm{s}},\\
\end{aligned}
\]
where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in
space $\bm{s}$ is equal to mean $\mu_s$. That mean is the result of an inverse
link function $g^{-1}$ applied to a linear predictor $\eta$. In this case, that
linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$
multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial
GRF $\omega_{\bm{s}}$. This spatial GRF represents the effect of latent spatial
variables that are not otherwise accounted for in the model. Alternatively,
$\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
More simply, $\omega_{\bm{s}}$ represents a ``wiggly'' surface with an expected
value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}).

The vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) is assumed drawn from
a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$ constrained by some
function that defines the rate that spatial covariance decays with
distance,
\[
\boldsymbol{\omega} \sim \operatorname{MVNormal}
  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right).
\]
Various covariance functions are possible, but a popular and flexible one is the
Mat\'ern \citep{matern1960} (Figure~\ref{fig:matern-range}). If we define $\lVert h \rVert$ as the Euclidean
distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the
Mat\'ern covariance $\Phi$ as
\[
\Phi\left( \bm{s}_j, \bm{s}_k \right) =
  \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \right).
\]
The parameter $\tau^2$ controls the spatial variance, $\Gamma$ represents the
Gamma function, $K_\nu$ represents the modified Bessel function of the second
kind, and $\kappa$ represents the spatial decorrelation rate. The parameter
$\nu$ controls the smoothness of the covariance function, but can be fixed at
$\nu = 1$ to take advantage of the SPDE \citep{lindgren2011}. The parameters
$\kappa$ and $\tau$ can be converted to two useful values: (1) correlation range
as $\sqrt{8} / \kappa$ and (2) marginal standard deviation as
$1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
The range is the distance that correlation decays to $\approx 0.13$ (Figure~\ref{fig:matern-range}).
